Relative Stable Maps in Genus One via Central Alignments

### 2019-01-28 GenusOneRelativeMaps.tex

Now, the logarithmic divisors which are contained in $\Dcal(k)$ are precisely those whose corresponding combinatorial type has the vertex of the dual graph containing $x_k$ mapped into the interior of $\RR_{\geq 0}$.

Thus, via the above procedure, we are able to enumerate the irreducible components of $\Dcal(k)$ in a combinatorial manner. The combinatorial type and choice of ray allows us to describe the general element of such a component. What we are still lacking is a recursive description of the components which will allow us to compute integrals over them. This is the subject of the remainder of this section.

@@

Now, let $\Dcal_{\Mcal}^\circ\subseteq \Mcal$ denote the (codimension-$1$) locally closed stratum corresponding to our choice of combinatorial type, and let $\Dcal_\Mcal = \overline{\Dcal^\circ_\Mcal}$ be the corresponding logarithmic stratum in $\Mcal$. Our divisor $\Dcal\subseteq\VZ$ is the proper transform of $\Dcal_\Mcal$, which is obtained by taking the preimage $\Dcal^\circ$ of $\Dcal_\Mcal^\circ$ (since $\Dcal^\circ_\Mcal$ is disjoint from the blown-up locus) and then taking its closure in $\VZ$. This is, of course, different from taking the preimage of $\Dcal_\Mcal$, since in general $\Dcal_\Mcal$ will intersect the blown-up locus. By continuity we have an induced map:
\begin{equation*} \Dcal \to \Dcal_\Mcal. \end{equation*}
The remainder of this section will be dedicated to showing how the map $\Dcal \to \Dcal_\Mcal$ may be interpreted as a blow-up. The reason this is useful is that $\Dcal_\Mcal$ itself has a recursive description in terms of moduli spaces of punctured maps (see \S \ref{subsection punctured maps}), which in our setting we are able to compute integrals over.

For the remainder of this subsection, we will consider only the situation where the circuit has positive degree; the other, more complicated case, will be taken up in \S \ref{}.

@@

\noindent The terminology is due to Vakil, and the picture in this case is very similar to \cite{Vre} (as we will see later, when the circuit is contracted the picture becomes \emph{very} different). Note that in these pictures we have omitted the marked points (apart from $x_k$), the degree of each vertex, and the expansion factors of the edges. These combinatorial data can be distributed arbitrarily, as long as:
\begin{enumerate}
\item every vertex has positive degree;
\item the balancing condition is satisfied.
\end{enumerate}
Note that since every vertex has positive degree, there is no need to impose stability.\end{lemma}

\begin{proof}
Recall that we have $\Dcal\subseteq\VZ$ obtained by a choice of combinatorial type and ray in the subdivison of the tropical moduli space. In this situation, the subdivision is trivial since no alignment is necessary (both radii are zero). Following our procedure, we choose a ray of the tropical moduli space $Q^\vee_{\RR}$. Since this is generated by edge lengths, and we are assuming that the initial and generised combinatorial types coincide, we conclude that the tropical moduli space must be $\RR_{\geq 0}$. Since we are assuming that $\Dcal \subseteq \Dcal(k)$, we know that the vertex of the tropicalisation which contains the flag $x_k$ must map into the interior $\RR_{>0}\subseteq \RR_{\geq 0}$.

We claim that this is the only vertex of the tropical curve mapped into the interior; if there were more, the positions of their images in $\RR_{\geq 0}$ would be independent, and thus the tropical moduli space would have dimension $\geq 2$, a contradiction.\footnote{(Luca) Say something about rigid tropical curves as in ACGS and KLR}

Thus there is a single vertex $\sqC_0$ mapped into the interior of $\RR_{\geq 0}$ and a number of vertices $\sqC_1,\ldots,\sqC_r$ mapped onto the vertex $0 \in \RR_{\geq 0}$. No pair of $\sqC_1,\ldots,\sqC_r$ can be connected by an edge, since this would introduce an extra edge length to the tropical moduli space, which then once again would have dimension $\geq 2$. Each of $\sqC_1,\ldots,\sqC_r$ must have positive degree by the tropical balancing condition and must be connected to $\sqC_0$ (since the tropical curve must be connected).

@@

\subsubsection{Type $A$} Let $\Delta$ be a combinatorial type of type $A$ and let $\Dcal\subseteq\VZ$ be the corresponding logarithmic divisor (note there is no choice of ray here). The corresponding locus $\Dcal_\Mcal\subseteq\Mcal$ is given (up to a finite cover) by\footnote{(Navid) Include short appendix on integralisation/saturation when gluing punctured maps in our setting?}:
\begin{equation*}\Dcal_\Mcal=\Mpunct{0}{\alpha^{(0)}\cup (-m_1,\ldots,-m_r)}{H}{d_0}\times_{H^r}\left( \Mlog{1}{\alpha^{(1)}\cup(m_1)}{\PP^N|H}{d_1} \times \prod_{i=2}^r\Mlog{0}{\alpha^{(i)}\cup(m_i)}{\PP^N|H}{d_i} \right). \end{equation*}
In this subsection we will show that
\begin{equation}\label{type A splitting axiom}\Dcal=\Mpunct{0}{\alpha^{(0)}\cup (-m_1,\ldots,-m_r)}{H}{d_0}\times_{H^r}\left(\VZ_{1,\alpha^{(1)}\cup(m_1)}(\PP^N|H,d_1)\times\prod_{i=2}^r\Mlog{0}{\alpha^{(i)}\cup(m_i)}{\PP^N|H}{d_i}\right) \end{equation}
i.e. that the map $\Dcal\to\Dcal_\Mcal$ is given by blowing up one of the factors of the fibre product. Note that the map $\Dcal\to\Dcal_\Mcal$ is an isomorphism away from the blown-up locus, which lives inside the locus where the circuit is contracted. Note that there is a birational map which forgets the log structures
\begin{equation*} \Mpunct{0}{\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}{H}{d_0} \to \M{0}{n_0+r}{H}{d_0} \end{equation*}
and all of our insertions are pulled back from the latter space. Therefore the integrals over the punctured space are determined by the genus zero Gromov--Witten theory of $H\cong\PP^{N-1}$.

@@

%Note that if $\sqC_0$ degenerates further then it may be the case that there are components of $\sqC_0$ contained inside the contraction radius for $\sqC$, and thus once we contract to a Smyth curve there are components of $\sqC_0$ adjacent to the genus one singularity. However, in this case when we split off $\sqC_1$ these components disappear, and if the factorisation condition holds for $\sqC$ then it holds a fortiori for $\sqC_1$.\footnote{(Navid) I think this is right, but could someone check? Maybe there's a better way to say this.} Thus we obtain a morphism from $\Dcal$ to the fibre product on the right-hand side of \eqref{type A splitting axiom}.

Given an element of $\Dcal$ we can split it along the nodes $q_1,\ldots,q_r$ as in \cite{PuncturedMaps}. It is then clear that $\sqC_1$ is aligned. We claim that $\sqC$ satisfies the factorisation property if and only if $\sqC_1$ does. This is enough to show \eqref{type A splitting axiom}.

@@

\begin{proof} The basic point is that $\Dcal$ consists of the union of the locally closed logarithmic strata adjacent to the locally closed stratum where all the $\sqC_i$ are irreducible. If we look at one of these boundary strata, the tropical moduli space contains a ray $\sigma$ corresponding to the stratum where all the $\sqC_i$ are irreducible; this amounts to setting all edge lengths other than $e_1,\ldots,e_r$ to zero. If $f_1,\ldots,f_l$ are some number of these additional edge lengths (corresponding to internal nodes in degenerations of the $\sqC_i$) then all of the cones of the subdivision adjacent to $\sigma$ will have $f_1+\ldots+f_l < e_j$ for all $j\in\{1,\ldots,r\}$, since $f_1=\ldots=f_l=0$ and $e_j \neq 0$ on $\sigma$. In particular, if $\sqC_1$ is degenerate and if $f$ denotes the minimal distance from the circuit to a non-contracted vertex of $\sqC_1$ (which certainly exists since $\sqC_1$ has positive degree) then $f < e_1 \leq \lambda(\sqC^\prime)$. Thus $\delta=f$ and $\delta < \lambda(\sqC^\prime)$ as claimed.\footnote{(Navid) Do an example?}\end{proof}
Notice we did not specify whether $\delta$ was the relative or absolute radius. The point is that it does not matter; the proof goes through the same in either case. Thus we see that $\sqC$ satisfies the (double) factorisation condition if and only if $\sqC_1$ does.

>>>>

Now, the logarithmic divisors which are contained in $\Dcal(k)$ are precisely those whose corresponding combinatorial type has the vertex of the dual graph containing $x_k$ mapped into the interior $\RR_{>0} \subseteq \RR_{\geq 0}$.

Thus, via the above procedure, we are able to enumerate the irreducible components of $\Dcal(k)$ in a combinatorial manner. The combinatorial type and choice of ray allows us to describe the general element of such a component. What we are still lacking is a recursive description of the components, which we need in order to compute integrals over them. This is the subject of the remainder of this section.

@@

Since $\VZ \to \Mcal$ is a log modification, the divisor $\Dcal$ is either exceptional or the proper transform of a logarithmic divisor on $\Mcal$. In this subsection, we will focus on the latter case. Then if we let $\Dcal_{\Mcal}^\circ\subseteq \Mcal$ denote the locally closed stratum corresponding to our choice of combinatorial type, and $\Dcal_\Mcal = \overline{\Dcal^\circ_\Mcal}$ denote the corresponding logarithmic divisor in $\Mcal$, then our divisor $\Dcal\subseteq\VZ$ is the proper transform of $\Dcal_\Mcal$, which is obtained by taking the preimage $\Dcal^\circ$ of $\Dcal_\Mcal^\circ$ (since $\Dcal^\circ_\Mcal$ is disjoint from the blown-up locus) and then taking its closure in $\VZ$. This is, of course, different from taking the preimage of $\Dcal_\Mcal$, since in general $\Dcal_\Mcal$ will intersect the blown-up locus. By continuity we have an induced map:
\begin{equation*} \Dcal \to \Dcal_\Mcal. \end{equation*}
The remainder of this section will be dedicated to showing how the map $\Dcal \to \Dcal_\Mcal$ may be interpreted as a blow-up. The reason this is useful is that $\Dcal_\Mcal$ itself has a recursive description in terms of moduli spaces of punctured maps (see \S \ref{subsection punctured maps}), which in our setting we are able to compute integrals over.

For the remainder of this subsection, we will consider only the situation where the circuit is assigned a positive degree by the combinatorial type; the other, more complicated case, will be taken up in \S \ref{}.

@@

\noindent The terminology is due to Vakil, and the picture in this case is very similar to \cite{Vre} (as we will see later, when the circuit is contracted the picture becomes \emph{very} different). Note that in these pictures we have omitted the marked points (apart from $x_k$), the degree of each vertex, and the expansion factors of the edges. These combinatorial data can be distributed arbitrarily, as long as:
\begin{enumerate}
\item the vertices $\sqC_1,\ldots,\sqC_r$ have positive degree (and $\sqC_0$ has positive degree in the $\Ycal_C^+$ case);
\item every vertex is stable;
\item the balancing condition is satisfied.
\end{enumerate}\end{lemma}

\begin{proof}
Recall that we have $\Dcal\subseteq\VZ$ obtained by a choice of combinatorial type and ray in the subdivison of the tropical moduli space. In this situation, since the circuit has positive degree no alignment is necessary (both radii are zero), and hence the subdivision is trivial. Following our procedure, we choose a ray of the tropical moduli space $Q^\vee_{\RR}$. Since this is generated by edge lengths, and we are assuming that the initial and generised combinatorial types coincide, we conclude that the tropical moduli space must be $\RR_{\geq 0}$. Since we are assuming that $\Dcal \subseteq \Dcal(k)$, we know that the vertex of the tropicalisation which contains the flag $x_k$ must map into the interior $\RR_{>0}\subseteq \RR_{\geq 0}$.

We claim that this is the only vertex of the tropical curve mapped into the interior; if there were more, the positions of their images in $\RR_{> 0}$ would be independent, and thus the tropical moduli space would have dimension $\geq 2$, a contradiction.\footnote{(Luca) Say something about rigid tropical curves as in ACGS and KLR}

Thus there is a single vertex $\sqC_0$ mapped into the interior $\RR_{>0} \subseteq \RR_{\geq 0}$ and a number of vertices $\sqC_1,\ldots,\sqC_r$ mapped onto the vertex $0 \in \RR_{\geq 0}$. No pair of $\sqC_1,\ldots,\sqC_r$ can be connected by an edge, since this would introduce an extra edge length to the tropical moduli space, which then once again would have dimension $\geq 2$. Each of $\sqC_1,\ldots,\sqC_r$ must have positive degree by the tropical balancing condition and must be connected to $\sqC_0$ (since the tropical curve must be connected).

@@

\subsubsection{Type $A$} Let $\Delta$ be a combinatorial type of type $A$ and let $\Dcal\subseteq\VZ$ be the corresponding logarithmic divisor (note there is no choice of ray here). The corresponding locus $\Dcal_\Mcal\subseteq\Mcal$ is given (up to a finite cover) by\footnote{(Navid) Include short appendix on integralisation/saturation when gluing punctured maps in our setting?}:
\begin{equation*}\Dcal_\Mcal=\Mpunct{0}{\alpha^{(0)}\cup (-m_1,\ldots,-m_r)}{H}{d_0}\times_{H^r}\left( \Mlog{1}{\alpha^{(1)}\cup(m_1)}{\PP^N|H}{d_1} \times \prod_{i=2}^r\Mlog{0}{\alpha^{(i)}\cup(m_i)}{\PP^N|H}{d_i} \right). \end{equation*}
\begin{lemma}We have the following description of $\Dcal$\begin{equation}\label{type A splitting axiom}\Dcal=\Mpunct{0}{\alpha^{(0)}\cup (-m_1,\ldots,-m_r)}{H}{d_0}\times_{H^r}\left(\VZ_{1,\alpha^{(1)}\cup(m_1)}(\PP^N|H,d_1)\times\prod_{i=2}^r\Mlog{0}{\alpha^{(i)}\cup(m_i)}{\PP^N|H}{d_i}\right) \end{equation}
i.e. the map $\Dcal\to\Dcal_\Mcal$ is given by blowing up one of the factors of the fibre product.\end{lemma} Note that there is a birational map which forgets the log structures
\begin{equation*} \Mpunct{0}{\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}{H}{d_0} \to \M{0}{n_0+r}{H}{d_0} \end{equation*}
and all of our insertions are pulled back from the latter space. Therefore the integrals over the punctured space are determined by the genus zero Gromov--Witten theory of $H\cong\PP^{N-1}$.

@@

%Note that if $\sqC_0$ degenerates further then it may be the case that there are components of $\sqC_0$ contained inside the contraction radius for $\sqC$, and thus once we contract to a Smyth curve there are components of $\sqC_0$ adjacent to the genus one singularity. However, in this case when we split off $\sqC_1$ these components disappear, and if the factorisation condition holds for $\sqC$ then it holds a fortiori for $\sqC_1$.\footnote{(Navid) I think this is right, but could someone check? Maybe there's a better way to say this.} Thus we obtain a morphism from $\Dcal$ to the fibre product on the right-hand side of \eqref{type A splitting axiom}.
\begin{proof}
Note first of all that the map $\Dcal\to\Dcal_\Mcal$ is an isomorphism away from the blown-up locus, which is contained inside the locus where the circuit is contracted. Given an element of $\Dcal$ we can split it along the nodes $q_1,\ldots,q_r$ as in \cite{PuncturedMaps}. It is then clear that $\sqC_1$ is aligned. We claim that $\sqC$ satisfies the factorisation property if and only if $\sqC_1$ does. This is enough to show \eqref{type A splitting axiom}.

@@

\begin{proof} The basic point is that $\Dcal$ consists of the union of the locally closed logarithmic strata adjacent to the locally closed stratum where all the $\sqC_i$ are irreducible. If we look at one of these boundary strata, the tropical moduli space contains a ray $\sigma$ corresponding to the stratum where all the $\sqC_i$ are irreducible; this amounts to setting all edge lengths other than $e_1,\ldots,e_r$ to zero. If $f_1,\ldots,f_l$ are some number of these additional edge lengths (corresponding to internal nodes in degenerations of the $\sqC_i$) then all of the cones of the subdivision adjacent to $\sigma$ will have $f_1+\ldots+f_l < e_j$ for all $j\in\{1,\ldots,r\}$, since $f_1=\ldots=f_l=0$ and $e_j \neq 0$ on $\sigma$. In particular, if $\sqC_1$ is degenerate and if $f$ denotes the minimal distance from the circuit to a non-contracted vertex of $\sqC_1$ (which certainly exists since $\sqC_1$ has positive degree) then $f < e_1 \leq \lambda(\sqC^\prime)$. Thus $\delta=f$ and $\delta < \lambda(\sqC^\prime)$ as claimed.\footnote{(Navid) Do an example?}\end{proof}
\noindent Notice we did not specify whether $\delta$ was the relative or absolute radius. The point is that it does not matter; the proof goes through the same in either case. Thus we see that $\sqC$ satisfies the (double) factorisation condition if and only if $\sqC_1$ does.\end{proof}
